# Directories and data frames - Week 4A - Workshop 1

This week we don't have any lectures but we do have an in person workshop. The section to complete this week may be a little more chunky then unusual, complete what you can during the workshop and then finish it off in your own time. Do make use of the demonstrators if any of the exercises so far haven't made sense. 

## Data Entry

Go to Blackboard > Skills for Biologists > Data Science > Workshop 1 and download the .jpeg files you find there. 

__List files here __

These are image files of lab notes taken during the data collection phase of a small research project to investigate parasite load in 2 small populations of the flour beetle _Tribolium castaneum_. Each beetle has an ID, a measure of wing casing in mm (which is a good proxy for overall size) and a tally for the number of tapeworm cysticercoids dissected out. In terms of data handling and analysis, we need this data to be stored digitally and we need to have it ordered sensibly in a spreadsheet. We covered what to look for in a tidy, well ordered spreadsheet in week 3 lectures. Have a go at entering the data from the .jpeg files into an Excel spreadsheet. When you're done go to __File > Save As...__, name your spreadsheet something sensible and make sure you select a sensible place to save it to, where it says __File Format:__ select __Comma-separated Values (.csv)__. 

## A note on naming

While we are on the subject of naming, I am going to take a moment to go over a few good rules to follow. You have just created names for some variables and for a file. You will also frequently have to name folders/directories and objects while working in R, so naming is something that you are repeatedly going to have to do. It therefore makes sense for you to learn how to do it well. 

* Keep names meaningful - Over the course of your degree you will create a lot of files and directories. You will also frequently be naming data variables and objects. You need to be able to easily recognise what is what so informative names are better then arbitrary names. So `girraffe`, `elephant`, `hippo` are better names then `1`, `2` and `3`. 

* Make versions obvious - There will be occasions when you want to save new information in a file or object but keep the original untouched. You may be experimenting or you may want the option to revisit to your original work. In these instances it makes sense to save your edits to a new version of the original file or object. If you do this and especially if you are sharing it with others make it clear. I tend to add my initials and a version identifier (e.g. v1, v2) as a suffix to the file name. 

* Keep names short - Typing takes time, its boring, and the more you type the greater the risk of a typo. So keep names short. So `trunk_length` is better than `elephant_trunk_length_cm`, there is important information in the second name but you can store this as a comment or a note. 

* Avoid spaces - Lots of software packages hate spaces, including R. If you can get out of the habbit of using spaces and instead use `_` your future self will thank you. 

* Abide by the rules for R - R does have some rules of its own for naming. Names must __not__ start with a number, but numbers may be used provided the first character is a letter. You can also use `.` or `_`, but avoid spaces at all costs. You may use upper or lower case letters, but remember R is case sensitive so will see `Elephant` and `elephant` as different. I try to keep to a lower case and use `_` instead of spaces, it keeps everything uniform and saves me time (this style is known as snake_case).

Now go back to your spreadsheet, do your variables and file names follow a sensible naming convention?

__Make sure you have a demonstrator check your work before moving onto the next step.__

## Setting up your R workspace 

This is a new piece of analysis, we will be carrying forward principals from the previous chapters but scripts and data will be new. So its a good idea to start with a fresh clean workspace. So lets set up a new project in RStudio Cloud, follow the steps from Chapter 2 if you cant remember how to do this and name this project `tribolium_parasites`. Its always good to keep a well organised tidy workspace, generally I will start a new project in RStudio Cloud if I am working on a new series of questions with a new data set or sets.

For this project you need to make sure two packages are installed. Copy the following directly into your console. 

```{r}
install.packages("tidyverse")
install.packages("janitor")
```


We will be moving towards doing some more comprehensive data handling so setting up your workspace is going to become more and more important. If you can get into the habit of setting up your workspace properly at the start of each project you will thank yourself later. A well developed project will contain several files including;

* Input data 
* Scripts
* Outputs

Under __Files__ in the bottom right panel create a __New Folder__, for each of the following;

* data
* scripts
* figures

__Make sure you copy the names over exactly, or some of the commands that we will be using later wont work__

A tidy workspace will help you find the things you need when you need them and will reduce the risk of loosing things. 

## A note on directories and their structure 

Working with text based interfaces like R, where you are writing and running commands, means that we need to have some understanding of the underlying computer architecture. You need to have some appreciation for how files are organised within folders/directories and why its important to know which directory you are working from with R.

Under Files in the bottom right panel of your screen you will see a file with the ending .Rproj. This is an R project file which tells R where you are working in the server. It means that R will automatically treat your project location as the __working directory__. If you wish to access files in any of the folders you have just created you will need to tell R the __path__ that it needs to follow in order to access these files. There are two types of path;

* __absolute__ paths, which refer to the same location in a file system relative to the root directory (don't worry if that doesn't make too much sense at this stage).
* __relative__ paths, which refer to a location in the file system relative to the directory you are currently working in.

Here we will only be dealing with __relative__ paths because you will always be working from within your R project directory. So for example if you wish to access a file in your `data` folder in the R project you would need to provide the path `("data/any_sub_folders_you_may_have_made/the_file_you_wish_to_access")`. We will do this in the next step so don't worry if you don't quite get it yet. 

## Setting up your script

Right OK, enough theory, lets start setting up your script. Create a new R script and call it `tribolium_parasites`. Its always good to add a few lines of description at the start of your script, this tells your future self and anyone else what the script is doing. Copy over the following.

```{r}
# An analysis of parasite abundance between male and female tribolium beetles from two thermal treatments (30 and 35 degrees centigrade)
# Your name
```

Then we need to load any packages that we might want to use;

```{r}
library(tidyverse) # Loads the tidyverse package
library(janitor) # Loads the janitor package
```

## Importing data

Lets import your data set. Go into your __data__ folder, click __Upload__ and select your newly made `.csv` file. Once its uploaded you should be able to see it under `data/` in the bottom right panel. 

Now we need to tell R to read your `.csv` file.

```{r}
# Read the data from the .csv file in the data folder into my project
tribolium_parasites <- read_csv("data/your_file_name.csv") #
```

R should respond with this;

```
> Tribolium <- read_csv("data/Tribolium_parasites.csv")
Rows: 20 Columns: 3                                                              
── Column specification ───────────────────────────────────────────────────────────
Delimiter: ","
dbl (3): beetle_id, wing_case_size, parasite_burden
```

Here R has told us that there are 20 rows and 3 columns in our data set, it has indicated that it has used comas as a collumn deliminator and that all the data are dbl (short for double), whidch means each collumn contains numbers. You should also be able to see `tribolium` in your __Environment__ under __Data__. 

If you have errors, make sure your file name and path are input correctly.

Lets have a look at our data to make sure it looks as we expect it to, try this;

```{r}
head(tribolium_parasites)
```

We can see that R has read from this we can see for ourselves that R has correclty read our data.

## Creating tibbles

So we have successfully imported data into R from a `.csv` file and stored it as a data frame which is a type of two dimensional array. Tibbles are a type of data frame as well, these can be useful if you want to type your data straight into R, or if you have more information to add to a pre-existing data frame. For example, we also have some additional data on the sex and treatment conditions of the Tribolium beetles. So lets create a tibble, add the following to your script;

```{r}
# First of all lets create some variables
beetle_id <- c(1:20) # Notice that this notation will create of a list of numbers 1 to 20
sex <- c("F","F","F","F","F","M","M","M","M","M","F","F","F","F","F","M","M","M","M","M")
treatment <- c(35,35,35,35,35,35,35,35,35,35,30,30,30,30,30,30,30,30,30,30)
```

Now we neet to turn our new variables into a data frame

```{r}
# Turning variables into a tibble
tribolium_extras <- tibble(beetle_id,sex,treatment)
```

Try using `head()` to check your new `tribolium_extras' data frame. 

It may be advantageous to have all the data from both of our data frames together, we can use another function to do this. However first I would like everyone in the class to have the same variable names, this will help us later on but also sometimes its useful to change headings in R. Add the following to your script and run it, make sure you edit it to match your variable names. 

```{r}
# Renaming some variables in your original data set
tribolium <- tribolium %>%
  rename(beetle_id = your_variable_name_here)
# Check how your data frame now looks
head(tribolium) 
```

Side note - you may have noticed the %>% expression, this is known as the forward pipe operator and allows you to chain together commands. This is a very simple use of it but it is piping your data frame forward to the rename function.

Hopefully the column containing your beetle IDs is now labeled `beetle_id`. Use the same set of commands to rename your variables for beetle wing case size and parasite load. Call these columns `wing_case_size` and `parasite_burden` instead. 

Now we can merge our two data frames. 

```{r}
# Merge data frames
tribolium_combo <- merge(tribolium, tribolium_extras, by = "beetle_id")
# This merge function combines both of your data frames but matches rows by beetle_id
```

Take a look at your new data frame now with the `head()` function. Lets try one more command to do one last check on your data. Try this;

```{r}
glimpse(tribolium_combo)
```

This command will show you something like this 

```
>glimpse(tribolium_combo)
Rows: 20
Columns: 5
$ beetle_id       <dbl> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17…
$ wing_case_size  <dbl> 2.4, 2.2, 2.5, 2.3, 2.3, 2.6, 2.6, 2.3, 2.4, 2.4, 2.3, 2.…
$ parasite_burden <dbl> 15, 15, 19, 13, 12, 19, 22, 10, 14, 16, 11, 8, 15, 9, 14,…
$ sex             <chr> "F", "F", "F", "F", "F", "M", "M", "M", "M", "M", "F", "F…
$ treatment       <dbl> 35, 35, 35, 35, 35, 35, 35, 35, 35, 35, 30, 30, 30, 30, 3…
```

Once again we can see that four of our variables have been interpreted by R as being `<dbl>` vectors, R has correctly identified them as being numeric. One of our variables has been identified as being a `<chr>` (short for character) vector, so R knows that this variable contains letters or words. 

You may now wish to export your new combined data frame as a `.csv` file. You can do this by running the following command;

```{r}
# Save your combined data frame as a .csv file (which you could download) to your data folder
write.csv(tribolium_combo,"data/tribolium_combo.csv", row.names = FALSE)
```

Well done everyone, you have successfully entered your data and uploaded it to RStudioCloud as well as making some small adjustments to the data frame and doing a first pass quality check! There are more checks we could have run, and we will revisit this later in the course when we look at some larger data sets. But we will leave this here for now, next week is reading week and the week after we will start looking at some initial observations in the data. 


## Before you leave!

Make sure you save your script!

### References

Firke, Sam. 2021. Janitor: Simple Tools for Examining and Cleaning Dirty Data. https://github.com/sfirke/janitor.
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D’Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. “Welcome to the tidyverse.” Journal of Open Source Software 4 (43): 1686. https://doi.org/10.21105/joss.01686.
Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2021. Dplyr: A Grammar of Data Manipulation. https://CRAN.R-project.org/package=dplyr.
Wickham, Hadley, Jim Hester, and Jennifer Bryan. 2021. Readr: Read Rectangular Text Data. https://CRAN.R-project.org/package=readr.