# A plot to choose from - Week 7A

You should aim to complete this chapter in Week 7, Semester 1

## Relationships, differences and distributions

As discussed in lectures, the plot you choose to produce should be influenced by the question you are asking.

If you are;

* Looking at differences between variables - make something like a box plot or bar chart
* Looking at relationships between variables - make a scatter plot
* Looking at the distribution of data within a variable - make a histogram

## My first histogram in R 

Lets have a go at making a histogram. Log-in to your posit Cloud account and go to your `tribolium_parasites` project. 

Add the following to your script;

```{r}
# Making a histogram and saving it in an object
# Call the ggplot function and direct it to your data and define your x axis
size_histogram <- ggplot(data = tribolium_combo,aes(x = wing_case_size)) + 
  geom_histogram(bins = 6) # Tell ggplot that you want it to build a histogram with 6 equal sized bins
print(size_histogram) # Print your new figure
```

Your plot should appear in the plot tab in the bottom right panel of posit Cloud. 

A side note on histograms and bins. You may have noticed that we have only specified our x axis variable. This is because R is essentially calculating a frequency for the y axis. How many beetles have a 2.1mm long wing case? How many beetles have a 2.4mm long wing case? This what the y axis is based on. We have asked R to split our x axis variable into 6 equal parts (bins). Have a go at changing the number of bins, what happens to the histogram?

Now have a go at making a histogram for the `parasite_burden` variable. 

Have a look at both of your histograms, what do you think they are showing? We will come back to histograms and how to read them in lectures. 

## Exporting your histogram

You are going to want to save your plots, you can use the `ggsave` function to do this.

```{r}
# Saving outputs
ggsave("figures/tribolium_wincase_size_histogram.pdf", # Give R a path to save to and a file name
       plot = size_histogram, # Tell R what to save - in this case your object
       device = "pdf") # Tell R what file type to create, in this case a pdf
```

Note that if you remove the device line `ggsave` will automatically create a `.png` file. You can also use `width = , height = , units = "cm"` arguments to specify the size of your figure in cm. Try to incorporate these new arguments into the above chunk. Save both of your histogram outputs in your `figures` folder. 

## Before you leave!

Log out of posit Cloud and make sure you save your script!

## References

Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D’Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. “Welcome to the tidyverse.” Journal of Open Source Software 4 (43): 1686. https://doi.org/10.21105/joss.01686.
Wickham, Hadley, Winston Chang, Lionel Henry, Thomas Lin Pedersen, Kohske Takahashi, Claus Wilke, Kara Woo, Hiroaki Yutani, and Dewey Dunnington. 2021. Ggplot2: Create Elegant Data Visualisations Using the Grammar of Graphics. https://CRAN.R-project.org/package=ggplot2.